Memory-Efficient Mixed-Precision Implementations\\for Robust Explicit Model Predictive Control

### 2019-04-15 emsoft-2019/Cont_Syn.tex
small fixes
notation

 while true:
   F, G, H, K = design_robust_MPC(model, $\Delta$)
   errorRegion = incorrect_region_error(F, G, H, K, $\varepsilon$)
   if $\Delta$ > errorRegion:
     errorAct = $\Delta$ - errorRegion
     $\hat{\text{F}}$, $\hat{\text{G}}$ = precision_tuning(F, G, errorAct)
     $\hat{\text{H}}$, $\hat{\text{K}}$ = precision_tuning(H, K, $\varepsilon_Q$)
     return $\hat{\text{F}}$, $\hat{\text{G}}$, $\hat{\text{H}}$, $\hat{\text{K}}$
   else:
     $\Delta$ = errorRegion + $\varepsilon_{SAFE}$
\end{lstlisting}
\caption{Algorithm for designing memory-efficient robust EMPC controller}
\label{lst:alg}
\end{figure}

@@

Given a state-space model of a plant (line 1), our algorithm returns four matrices
$\hat{\text{F}}$, $\hat{\text{G}}$, $\hat{\text{H}}$, and $\hat{\text{K}}$,
which soundly implement a robust explicit MPC controller in mixed-precision
fixed-point arithmetic.
Our algorithm takes two additional inputs, $\Delta$ and $\varepsilon_Q$ (line 2, 3). 
The disturbance $\Delta$ provided by the user represents a starting point for 
the search for a robust controller. In principle, the user can set $\Delta = 0$,
however note that the finite-precision implementation of the controller will always
incur at least some small error. In practice we thus start the search with a slightly
larger value, e.g. $\Delta = 0.1$. Our algorithm will perform a linear search and
increase $\Delta$ if needed.

@@

We need to distribute the available `disturbance budget' among the two error
sources: error due to selecting the wrong region, which is given by the
quantization of $\hat{\text{H}}$ and $\hat{\text{K}}$, and the error due to
quantization of the control action, given by $\hat{\text{F}}$, $\hat{\text{G}}$.
The quantization of $\hat{\text{H}}$ and $\hat{\text{K}}$ is determined by
$\varepsilon_Q$, thus by providing this input value, the user effectively
chooses how much precision to allow for $\hat{\text{H}}$ and $\hat{\text{K}}$.
Note that this precision comes at the expense of the precision of
$\hat{\text{F}}$ and $\hat{\text{G}}$, i.e. the more accurately $\hat{\text{H}}$
and $\hat{\text{K}}$ are represented, the more approximate $\hat{\text{F}}$ and
$\hat{\text{G}}$ need to be and vice versa.
We note that one could straight-forwardly extend our algorithm to include an
outer loop which would explore different values for $\varepsilon_Q$. For our
experiments in~\autoref{sec:experiments}, we vary this value manually.

@@

The algorithm then calls Matlab's multi-parametric toolbox to design an explicit
MPC controller for the given state-space model which is robust to the initial
disturbance $\Delta$ (line 8). The result is a controller represented by four matrices
\texttt{F, G, H, K} given in high precision.

\emph{Assuming} that we can implement the polyhedral region partitioning (given by
matrices \texttt{H, K}) with an implementation error of at most $\varepsilon_Q$,
our algorithm proceeds to compute the error due to selecting the wrong region (line 9).
This computation works as explained in~\autoref{sec:Error_Analysis}, \autoref{eq:maximization}.

@@

If the region error is smaller than $\Delta$, we still have an error budget \texttt{errorAct} left
over to implement the control actions in finite precision (line 11).
For the precision assignment, we first call Daisy with the expressions for the
control actions given by \texttt{F} and \texttt{G}, and the remaining error
budget as the error bound (line 12).
Then we call Daisy with the expressions of the region bounds given by \texttt{H}
and \texttt{K}, now with the error bound $\varepsilon_Q$ (line 13).

@@

We implemented our algorithm in a Python script which interfaces between Matlab
and Daisy and which implements the high-level structure from~\autoref{lst:alg}.
The interface parses the output from Matlab and encodes the matrices \texttt{F,
G, H, K} and the error bounds in Daisy's input format.

>>>>

 while true:
   $F$, $G$, $H$, $K$ = design_robust_MPC(model, $\Delta$)
   errorRegion = incorrect_region_error($F$, $G$, $H$, $K$, $\varepsilon$)
   if $\Delta$ > errorRegion:
     errorAct = $\Delta$ - errorRegion
     $\hat{F}$, $\hat{G}$ = precision_tuning($F$, $G$, errorAct)
     $\hat{H}$, $\hat{K}$ = precision_tuning($H$, $K$, $\varepsilon_Q$)
     return $\hat{F}$, $\hat{G}$, $\hat{H}$, $\hat{K}$
   else:
     $\Delta$ = errorRegion + $\varepsilon_{SAFE}$
\end{lstlisting}
\caption{Algorithm for designing memory-efficient robust EMPC controller}
\label{lst:alg}
\end{figure}

@@

Given a state-space model of a plant (line 1), our algorithm returns four matrices
$\hat{F}$, $\hat{G}$, $\hat{H}$, and $\hat{K}$,
which soundly implement a robust explicit MPC controller in mixed-precision
fixed-point arithmetic.
Our algorithm takes two additional inputs, $\Delta$ and $\varepsilon_Q$ (line 2, 3). 
The disturbance $\Delta$ provided by the user represents a starting point for 
the search for a robust controller. In principle, the user can set $\Delta = 0$,
however note that the finite-precision implementation of the controller will always
incur at least some small error. In practice we thus start the search with a slightly
larger value, e.g. $\Delta = 0.1$. Our algorithm will perform a linear search and
increase $\Delta$ if needed.

@@

We need to distribute the available `disturbance budget' among the two error
sources: error due to selecting the wrong region, which is given by the
quantization of $\hat{H}$ and $\hat{K}$, and the error due to
quantization of the control action, given by $\hat{F}$, $\hat{G}$.
The quantization of $\hat{H}$ and $\hat{K}$ is determined by
$\varepsilon_Q$, thus by providing this input value, the user effectively
chooses how much precision to allow for $\hat{H}$ and $\hat{K}$.
Note that this precision comes at the expense of the precision of
$\hat{F}$ and $\hat{G}$, i.e. the more accurately $\hat{H}$
and $\hat{K}$ are represented, the more approximate $\hat{F}$ and
$\hat{G}$ need to be and vice versa.
We note that one could straight-forwardly extend our algorithm to include an
outer loop which would explore different values for $\varepsilon_Q$. For our
experiments in~\autoref{sec:experiments}, we vary this value manually.

@@

The algorithm then calls Matlab's multi-parametric toolbox to design an explicit
MPC controller for the given state-space model which is robust to the initial
disturbance $\Delta$ (line 8). The result is a controller represented by four matrices
$F$, $G$, $H$, $K$ given in high precision.

\emph{Assuming} that we can implement the polyhedral region partitioning (given by
matrices $H$, $K$) with an implementation error of at most $\varepsilon_Q$,
our algorithm proceeds to compute the error due to selecting the wrong region (line 9).
This computation works as explained in~\autoref{sec:Error_Analysis}, \autoref{eq:maximization}.

@@

If the region error is smaller than $\Delta$, we still have an error budget \texttt{errorAct} left
over to implement the control actions in finite precision (line 11).
For the precision assignment, we first call Daisy with the expressions for the
control actions given by $F$ and $G$, and the remaining error
budget as the error bound (line 12).
Then we call Daisy with the expressions of the region bounds given by $H$
and $K$, now with the error bound $\varepsilon_Q$ (line 13).

@@

We implemented our algorithm in a Python script which interfaces between Matlab
and Daisy and which implements the high-level structure from~\autoref{lst:alg}.
The interface parses the output from Matlab and encodes the matrices $F$,
$G$, $H$, $K$ and the error bounds in Daisy's input format.

We set up the problem with YALMIP~\cite{matlabYALMIP}, and use 
Matlab's MPT toolbox~\cite{matlabMPT} to compute the robust explicit MPC.

### 2019-04-15 emsoft-2019/Methodology.tex
notation

The controller $u_{k}$ for (\ref{eq:statespacemodel}) is found by solving an off-line optimization problem. We rely on MATLAB multi-parametric toolbox to solve such optimization problem~\cite{matlabMPT, matlabYALMIP}. 
The output of the suite is a continuous piecewise affine (PWA) function, together with a partitioning of the domain of \statevarmath. Each element in the partitioning is called region.
A generic region $i$ consists in a n-polytope, uniquely identified by a set of inequalities in the form: $H_{i}\statevar<=K_{i}$.

We call \statespace\space the union of all regions $X_{i}$ in the domain of the state variable $x_{k}$:
\begin{equation}
\statespace = \bigcup_{i}(H_{i}\statevar<=K_{i})
\end{equation}

@@

A drawback in the design of explicit MPC is that both regions boundaries (H and K) and activation functions (F and G), have to be stored on the micro-controller. Usually these devices have limited memory, in the order of KBs.

@@

\subsection{Precision Tuning}
In formula (\ref{eq:epsilontot}), the equation for $\varepsilon$ contains: $\varepsilon_{A/D}$ that is the error introduced by the analog-digital conversion, and $\varepsilon_{Q}$ that is generated by the quantization of hyperplanes $H\statevar<=K$ in a finite number of bits.
This $\varepsilon_{Q}$ can be tuned based on memory availability in the micro-controller.
The rule for $\varepsilon_{Q}$ is the following:
\begin{equation}\label{eq:quantizationlines}
\varepsilon_{Q} > (H-\hat{H}_{p})\qstatevar+(K-\hat{K}_{p})
\end{equation}
This inequality assures that the distance between any hyperplane represented in infinite precision (H and K), and its counter-part quantized in p bits ($(\hat{H})_{p}$ and $(\hat{K})_{p}$), is bounded by $\varepsilon_{Q}$.

@@

Since the state variable \statevarmath is bounded by \statespace, and F,G,H, and K are vectors of constants, we can rely on Daisy for this verification step.

@@

while True:
design_robust_MPC(delta)
maxUij = compute (*(\ref{eq:maximization})*) with size(tube)=(*$\varepsilon$*)
if delta > maxUij:
max_err=delta-maxUij
UNI_MIX_precision(F,G,max_err)
UNI_MIX_precision(H,K,(*$\varepsilon_{Q}$*))
break
else:
delta=maxUij+(*$\varepsilon_{SAFE}$*)
\end{lstlisting}
In Listing \ref{lst:alg} we describe the procedure used to design a robust MPC controller such that (\ref{eq:delta}) is respected.

>>>>

The controller $u_{k}$ for (\ref{eq:statespacemodel}) is found by solving an off-line optimization problem. We rely on MATLAB multi-parametric toolbox to solve such optimization problem~\cite{matlabMPT, matlabYALMIP}. 
The output of the suite is a continuous piecewise affine (PWA) function, together with a partitioning of the domain of \statevarmath. Each element in the partitioning is called region.
A generic region $i$ consists in a n-polytope, uniquely identified by a set of inequalities in the form: $H_{i}\statevar\le K_{i}$.

We call \statespace\space the union of all regions $X_{i}$ in the domain of the state variable $x_{k}$:
\begin{equation}
\statespace = \bigcup_{i}(H_{i}\statevar\le K_{i})
\end{equation}

@@

A drawback in the design of explicit MPC is that both regions boundaries ($H$ and $K$) and activation functions ($F$ and $G$), have to be stored on the micro-controller. Usually these devices have limited memory, in the order of KBs.

@@

\subsection{Precision Tuning}
In formula (\ref{eq:epsilontot}), the equation for $\varepsilon$ contains: $\varepsilon_{A/D}$ that is the error introduced by the analog-digital conversion, and $\varepsilon_{Q}$ that is generated by the quantization of hyperplanes $H\statevar\le K$ in a finite number of bits.
This $\varepsilon_{Q}$ can be tuned based on memory availability in the micro-controller.
The rule for $\varepsilon_{Q}$ is the following:
\begin{equation}\label{eq:quantizationlines}
\varepsilon_{Q} > (H-\hat{H}_{p})\qstatevar+(K-\hat{K}_{p})
\end{equation}
This inequality assures that the distance between any hyperplane represented in infinite precision ($H$ and $K$), and its counter-part quantized in p bits ($(\hat{H})_{p}$ and $(\hat{K})_{p}$), is bounded by $\varepsilon_{Q}$.

@@

Since the state variable \statevarmath is bounded by \statespace, and $F,G,H$ and $K$ are vectors of constants, we can rely on Daisy for this verification step.

@@

while True:
design_robust_MPC(delta)
maxUij = compute (*(\ref{eq:maximization})*) with size(tube)=(*$\varepsilon$*)
if delta > maxUij:
max_err=delta-maxUij
UNI_MIX_precision($F$,$G$,max_err)
UNI_MIX_precision($H$,$K$,(*$\varepsilon_{Q}$*))
break
else:
delta=maxUij+(*$\varepsilon_{SAFE}$*)
\end{lstlisting}
In Listing \ref{lst:alg} we describe the procedure used to design a robust MPC controller such that (\ref{eq:delta}) is respected.

### 2019-04-15 emsoft-2019/experiments.tex
notation

\autoref{tab:ipd} shows the total number of bits required to implement each
controller using different precision options: uniform 32 bit precision
(`Uni32'), minimal uniform precision (`Uni', chosen precision in parentheses)
and mixed precision (`Mix'). We split the memory requirement into the bits
required for storing F and G and H and K. We show the results for each benchmark
for ten different combinations of $\Delta$ and $\varepsilon_Q$, varying one while
keeping the other fixed.
For the pendulum, the execution time of the whole analysis is 10 minutes no matter the
initial values for $\Delta$ and $\varepsilon_Q$, while for the aircraft it is 36
minutes.

@@

%\eva{Fix Table 1 caption. It's not the difference, but savings. Difference is absolute.}
%\eva{Compute the actual averages for all F, G, H, K and not only approximations}
For the inverted pendulum, minimal uniform precision saves on average $52.6\%$ of memory
compared to a uniform 32 bit baseline overall, i.e. for F, G, H and K together. 
Mixed-precision further reduces the memory
requirement by $10.5\%$ on average with respect to the minimal uniform baseline (and 57.5\% w.r.t to 
uniform 32 bit baseline). 
\autoref{tab:ipd} shows a more detailed breakdown of the memory requirements and savings
between F and G and H and K.
The memory requirements for the storage of hyperplanes H and K depends only on
the size of the tubes $\varepsilon_Q$ so that memory requirements remain constant
for fixed $\varepsilon_Q$.

%\eva{Compute actual and overall averages here too}
For the aircraft example, minimal uniform precision saves on average $19.3\%$
overall w.r.t. a uniform 32 bit baseline, and mixed-precision saves an
additional $17.7\%$ w.r.t. minimal uniform precision (33.6\% w.r.t to uniform 32 bit). We observe higher relative
memory savings by mixed-precision for storing H and K than for the inverted pendulum example.

@@

%We noticed the following common behavior in the two benchmarks: 
As expected, when $\Delta$ decreases, the control actions (F, G) need to be implemented
more precisely and require more memory, because the space for approximation error
is reduced. Similarly, when the value of $\varepsilon_Q$ increases, the memory
requirements for H, K can be relaxed.

%We note two differences among these two benchmarks.
We note that for the aircraft example,
the precision for F and G is almost double with respect to the pendulum. This is
because the magnitude of F and G is on the order of $10^{3}$ while for the
pendulum it is on the order of several units. Note, however, that for F and G the memory gain
from minimal uniform to mixed precision (\%UvM) is only slightly less than the one for the pendulum.

@@

\autoref{tab:di} shows the results of this experiment.
We observe that our analysis scales up to 1829 controllers and 14632 hyperplanes in a few hours.  The computed minimal uniform precision saves on average $44.5\%$ of memory
compared to a uniform $32$ bit baseline overall (for F, G, H and K together). 
Moreover, mixed-precision reduces the memory requirement by an additional
$9.3\%$ on average with respect to the minimal uniform baseline (49.6\% w.r.t. uniform 32 bit).
We further observe that relative savings remain largely constant for different 
prediction horizons.
% \autoref{tab:di} shows the stability of memory tuning in Daisy is
% constant with respect to the number of controllers or hyperplanes. This is true
% for both F, G and H, K. Indeed, the memory gain from uniform in 32bits and
% minimal uniform (\%32vU) is constant among different input values m. The same
% holds for (\%UvM).

>>>>

\autoref{tab:ipd} shows the total number of bits required to implement each
controller using different precision options: uniform 32 bit precision
(`Uni32'), minimal uniform precision (`Uni', chosen precision in parentheses)
and mixed precision (`Mix'). We split the memory requirement into the bits
required for storing $F$, $G$, $H$ and $K$. We show the results for each benchmark
for ten different combinations of $\Delta$ and $\varepsilon_Q$, varying one while
keeping the other fixed.
For the pendulum, the execution time of the whole analysis is 10 minutes no matter the
initial values for $\Delta$ and $\varepsilon_Q$, while for the aircraft it is 36
minutes.

@@

%\eva{Fix Table 1 caption. It's not the difference, but savings. Difference is absolute.}
%\eva{Compute the actual averages for all F, G, H, K and not only approximations}
For the inverted pendulum, minimal uniform precision saves on average $52.6\%$ of memory
compared to a uniform 32 bit baseline overall, i.e. for $F$, $G$, $H$ and $K$ together. 
Mixed-precision further reduces the memory
requirement by $10.5\%$ on average with respect to the minimal uniform baseline (and 57.5\% w.r.t to 
uniform 32 bit baseline). 
\autoref{tab:ipd} shows a more detailed breakdown of the memory requirements and savings
between $F$, $G$, $H$ and $K$.
The memory requirements for the storage of hyperplanes $H$ and $K$ depends only on
the size of the tubes $\varepsilon_Q$ so that memory requirements remain constant
for fixed $\varepsilon_Q$.

%\eva{Compute actual and overall averages here too}
For the aircraft example, minimal uniform precision saves on average $19.3\%$
overall w.r.t. a uniform 32 bit baseline, and mixed-precision saves an
additional $17.7\%$ w.r.t. minimal uniform precision (33.6\% w.r.t to uniform 32 bit). We observe higher relative
memory savings by mixed-precision for storing $H$ and $K$ than for the inverted pendulum example.

@@

%We noticed the following common behavior in the two benchmarks: 
As expected, when $\Delta$ decreases, the control actions ($F$, $G$) need to be implemented
more precisely and require more memory, because the space for approximation error
is reduced. Similarly, when the value of $\varepsilon_Q$ increases, the memory
requirements for $H$ and $K$ can be relaxed.

%We note two differences among these two benchmarks.
We note that for the aircraft example,
the precision for $F$ and $G$ is almost double with respect to the pendulum. This is
because the magnitude of $F$ and $G$ is on the order of $10^{3}$ while for the
pendulum it is on the order of several units. Note, however, that for $F$ and $G$ the memory gain
from minimal uniform to mixed precision (\%UvM) is only slightly less than the one for the pendulum.

@@

\autoref{tab:di} shows the results of this experiment.
We observe that our analysis scales up to 1829 controllers and 14632 hyperplanes in a few hours.  The computed minimal uniform precision saves on average $44.5\%$ of memory
compared to a uniform $32$ bit baseline overall (for $F$, $G$, $H$ and $K$ together). 
Moreover, mixed-precision reduces the memory requirement by an additional
$9.3\%$ on average with respect to the minimal uniform baseline (49.6\% w.r.t. uniform 32 bit).
We further observe that relative savings remain largely constant for different 
prediction horizons.
% \autoref{tab:di} shows the stability of memory tuning in Daisy is
% constant with respect to the number of controllers or hyperplanes. This is true
% for both F, G and H, K. Indeed, the memory gain from uniform in 32bits and
% minimal uniform (\%32vU) is constant among different input values m. The same
% holds for (\%UvM).
